---
title: "bereavement(): Period Bereavement Estimation with DemoKin"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  dpi = 150,
  fig.width  = 10,
  fig.height = 5
)
options(scipen = 999)
```

`bereavement()` is an R function that estimates how many people in a population lose a close relative in a given year. It is designed as a companion to the [DemoKin](https://github.com/IvanWilli/DemoKin) package, which provides the underlying kinship model.

This document walks through the full workflow from loading demographic data to producing bereavement estimates and visualisations. Along the way it explains the demographic theory behind every step, so it can serve as both a user guide and a methods reference.

---

## Background

### What is period bereavement?

A **period bereavement estimate** answers the question: *In a given calendar year, how many people lost at least one relative of type k?* This is a cross-sectional measure — a snapshot for a specific year — not a life-course cumulative probability.

Formally, for a focal individual of age *a* in year *y*, define:

```
q0(a, y, k) = P(focal person loses ≥1 relative of type k during year y)
```

The number of bereaved people of age *a* in year *y* is then:

```
B(a, y, k) = q0(a, y, k) × N(a, y)
```

where `N(a, y)` is the number of people of age *a* alive in year *y*. Summing over all ages gives the total bereaved count for kin type *k* in year *y*.

### The kinship surface and DemoKin

To compute `q0`, we first need to know how many living relatives of each type a person of age *a* is expected to have in year *y* — and how old those relatives are. This is the **kinship surface**, a central concept in formal demography.

Caswell (2019, 2022) showed that the kinship surface can be derived from a matrix population model. Given age-specific survival probabilities and fertility rates, the model tracks expected kin counts through the recursive structure of family trees. For example: the expected number of living mothers aged *a'* for a focal person of age *a* in year *y* depends on fertility rates *a* years ago and survival rates in every year since.

The `DemoKin` package implements this model in R. Its `kin2sex()` function runs a **two-sex, time-varying** version of the model, incorporating separate male and female demographic rates that can change over time. The output is a data frame `kin_full` with one row per combination of:

| Column | Meaning |
|--------|---------|
| `year` | Calendar year |
| `age_focal` | Age of the focal person (0–100) |
| `kin` | Relationship type (see table below) |
| `age_kin` | Age of the relative (0–100) |
| `sex_kin` | Sex of the relative (`"f"` or `"m"`) |
| `living` | Expected number of living relatives with this profile |
| `dead` | Cumulative expected deaths of this kin type up to `year` |

Common kin type codes:

| Code | Relationship | Notes |
|------|-------------|-------|
| `m` | Parent (mother/father) | Expected count peaks near 1 at birth |
| `d` | Child | Grows as focal person has children |
| `gm` | Grandparent | |
| `gd` | Grandchild | |
| `s` / `os` / `ys` | Sibling (older/younger) | |
| `a` / `oa` / `ya` | Aunt or uncle | |
| `n` / `nos` / `nys` | Niece or nephew | |
| `c` / `coa` / `cya` | Cousin | |

Because `living` is derived from a deterministic matrix model, it is a **continuous-valued** expected count, not an integer. A focal person aged 30 might have an expected 0.73 grandmothers aged 68 alive — the average over the population, not a count for any specific individual.

### The product formula

Given the kinship surface, how do we compute `q0(a, y, k)`?

`bereavement()` uses the **product method** (also called the "Kike method"). It rests on two assumptions:

1. The deaths of different relatives during year *y* are **independent** given their age and sex.
2. The actual number of relatives at each age follows a **Poisson distribution** with mean equal to `living`.

Under these assumptions, the probability that a specific cohort of `living` relatives of age *a'* and sex *g* all survive year *y* is:

```
(1 - q(a', y, g)) ^ living(a, a', k, y, g)
```

where `q(a', y, g)` is the age-sex-year-specific probability of dying. The exponent is continuous because `living` is continuous; it is interpreted as `exp(living × log(1 - q))`, which is the probability generating function of a Poisson distribution evaluated at `(1 - q)`. This expression is **exact** (not an approximation) under the Poisson + independence assumptions.

Taking the product over all ages *a'* and both sexes *g*:

```
P(all kin of type k survive | focal age a, year y)
  = ∏_{a', g} (1 - q(a', y, g)) ^ living(a, a', k, y, g)
```

Therefore:

```
q0(a, y, k) = 1 - ∏_{a', g} (1 - q(a', y, g)) ^ living(a, a', k, y, g)
```

Ages with `living = 0` contribute `(1-q)^0 = 1` and drop out of the product automatically. This means the function only needs the full `living` surface and age-sex-specific `qx` values; it does not require any knowledge of the actual family structure of any individual.

> **The independence assumption matters.** In populations where death clusters within families (conflict, epidemic, famine), the product method overestimates the number of bereaved people, because it spreads losses evenly rather than concentrating them in fewer families. For populations with such clustering, a Negative Binomial extension exists (Alburez-Gutierrez et al. 2024) but is not yet implemented here.

---

## Setup

Install `DemoKin` from GitHub if needed, then load packages and source the function:

```{r packages}
# install.packages("remotes")
# remotes::install_github("IvanWilli/DemoKin")

library(DemoKin)
library(tidyverse)

source("bereavement.R")
```

---

## Step 1 — Load demographic data

`DemoKin` ships three built-in datasets for Sweden (ages 0–100, years 1900–2018):

- `swe_px`: female annual survival probabilities `p(a, t)` — the probability of surviving from age *a* to *a+1* in year *t*
- `swe_asfr`: female age-specific fertility rates — live births per woman per year
- `swe_pop`: observed female population counts

```{r load-data}
pf <- swe_px    # female survival:  101 ages × 119 years
ff <- swe_asfr  # female fertility: 101 ages × 119 years
```

`DemoKin`'s built-in data are female only. The two-sex model requires male rates as well. Following the DemoKin vignette (Caswell 2022), we construct synthetic male rates:

```{r male-rates}
n_ages <- nrow(pf)
n_yrs  <- ncol(pf)

# Male survival: raised to power 1.5 → uniformly lower than female (male excess mortality)
pm <- pf ^ 1.5
dimnames(pm) <- dimnames(pf)

# Male fertility: shift female ASFR 5 years to the right → later mean age at fathering
fm <- rbind(matrix(0, 5, n_yrs), ff[-((n_ages - 4):n_ages), ])
dimnames(fm) <- dimnames(ff)
```

Let's check what these rates look like for a single year:

```{r plot-rates, fig.height=4}
yr_show <- "1980"

bind_rows(
  tibble(age = 0:100, sex = "Female", component = "Survival probability",  value = pf[, yr_show]),
  tibble(age = 0:100, sex = "Male",   component = "Survival probability",  value = pm[, yr_show]),
  tibble(age = 0:100, sex = "Female", component = "Fertility rate (ASFR)", value = ff[, yr_show]),
  tibble(age = 0:100, sex = "Male",   component = "Fertility rate (ASFR)", value = fm[, yr_show])
) |>
  ggplot(aes(age, value, colour = sex)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ component, scales = "free_y") +
  scale_colour_manual(values = c(Female = "#c0392b", Male = "#2980b9")) +
  labs(
    title  = paste("Demographic rates used as model inputs — Sweden", yr_show),
    x = "Age", y = NULL, colour = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

Male survival is lower at all ages. Male fertility peaks about 5 years later than female fertility. These are the rates that drive the kinship model.

---

## Step 2 — Compute the kinship surface with `kin2sex()`

`kin2sex()` is the two-sex kinship model in `DemoKin`. It takes survival and fertility matrices for both sexes and returns, for every (focal age × year × kin type × kin age × kin sex) combination, the expected number of living and dead relatives.

Key arguments:

| Argument | Description |
|----------|------------|
| `pf`, `pm` | Female and male survival probability matrices (101 ages × *n* years) |
| `ff`, `fm` | Female and male fertility matrices (same dimensions) |
| `sex_focal` | Sex of the focal individual (`"f"` or `"m"`) |
| `time_invariant` | `FALSE` uses rates that change over time (time-varying model) |
| `birth_female` | Proportion of births that are female (typically `1/2.04 ≈ 0.49`) |
| `output_kin` | Which relationship types to compute (limits time and memory) |

The time-varying model (`time_invariant = FALSE`) is computationally intensive — it must propagate rates from every birth cohort's birth year to the present. Results are cached automatically on first run.

```{r kin2sex, eval=FALSE}
kin_out <- kin2sex(
  pf = pf, pm = pm,
  ff = ff, fm = fm,
  sex_focal      = "f",
  time_invariant = FALSE,
  birth_female   = 1 / 2.04,
  output_kin     = c("d", "gd", "gm", "m", "s")
)
kin_full <- kin_out$kin_full
```

```{r kin2sex-cache, echo=FALSE}
cache_file <- "cache/kin_full_swe_demo.rds"
if (file.exists(cache_file)) {
  kin_full <- readRDS(cache_file)
} else {
  kin_out <- kin2sex(
    pf = pf, pm = pm,
    ff = ff, fm = fm,
    sex_focal      = "f",
    time_invariant = FALSE,
    birth_female   = 1 / 2.04,
    output_kin     = c("d", "gd", "gm", "m", "s")
  )
  kin_full <- kin_out$kin_full
  dir.create("cache", showWarnings = FALSE)
  saveRDS(kin_full, cache_file)
}
```

```{r kin-dims}
nrow(kin_full)             # total rows
range(kin_full$year)       # years covered
sort(unique(kin_full$kin)) # kin types computed
```

Here is what the first few rows look like:

```{r kin-head}
kin_full |>
  filter(year == 1980, age_focal == 40, kin == "m") |>
  arrange(sex_kin, age_kin) |>
  filter(living > 0) |>
  head(10)
```

Each row describes one (age × sex) slice of a kin type for a particular focal person profile. The `living` column is the expected number of relatives with that exact age and sex still alive. The `dead` column is the cumulative deaths of that kin type up to that year (not period deaths — see the HANDOFF for details).

### Visualising the kinship surface

To build intuition, let's look at the expected number of living relatives by focal age for a single year:

```{r plot-kin-surface, fig.height=5}
kin_full |>
  filter(year == 1980) |>
  summarise(living = sum(living), .by = c(age_focal, kin)) |>
  rename_kin(sex = "f") |>
  ggplot(aes(age_focal, living)) +
  geom_line(colour = "#2c3e50", linewidth = 0.8) +
  facet_wrap(~ kin_label, scales = "free_y", nrow = 2) +
  labs(
    title   = "Expected number of living relatives by focal age — Sweden 1980",
    subtitle = "Two-sex time-varying model. Female focal. Living counts summed over kin age and sex.",
    x = "Age of focal woman", y = "Expected living relatives"
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.subtitle    = element_text(colour = "grey40", size = 8),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

This is the kinship surface that feeds into the bereavement calculation. For example, a 30-year-old Swedish woman in 1980 expected to have roughly one living parent, no grandchildren yet, several siblings, and a growing number of children.

---

## Step 3 — Compute the probability of dying (`qx`)

The bereavement model needs `qx`: the probability that a person aged *x* dies within the year. This is derived from the survival probabilities:

```
qx = 1 - px
```

**Terminal age problem.** `swe_px[age = 100] = 0` because everyone in the open age group (100+) is assumed to die within the year in this life table. That gives `qx = 1`, which collapses the product formula to zero for any focal person with living kin aged 100+, and triggers an error in `bereavement()`.

The fix is to ensure `qx < 1` at the terminal age. We do this by replacing `px = 0` with a small positive value before computing `qx`. In a formal analysis, the correct approach is to derive `qx` from the force of mortality `mx` using `DemoTools::lt_single_mx()`, which handles the open age group properly.

```{r compute-qx}
# Replace any px = 0 with 1e-6 before computing qx, to avoid qx = 1 at age 100
qx_f <- 1 - pmax(pf, 1e-6)
qx_m <- 1 - pmax(pm, 1e-6)

dimnames(qx_f) <- dimnames(pf)
dimnames(qx_m) <- dimnames(pm)

range(qx_f)  # female qx: min and max
range(qx_m)  # male   qx: min and max
```

---

## Step 4 — Population data

`swe_pop` contains observed female population counts for Sweden by age and year. This is the population of focal women — we need it to convert per-capita probabilities into head counts:

```
B(a, y, k) = q0(a, y, k) × N(a, y)
```

```{r pop-data}
pop_mat <- swe_pop
dimnames(pop_mat) <- list(as.character(0:100), colnames(swe_pop))

# Total Swedish women in selected years
colSums(pop_mat)[as.character(c(1950, 1970, 1990, 2010))] |>
  formatC(format = "d", big.mark = ",") |>
  print()
```

---

## Step 5 — Run `bereavement()`

`bereavement()` applies the product formula to every combination of (year, focal age, kin type) in `kin_full` and returns a tidy data frame.

**Arguments:**

| Argument | Description |
|----------|------------|
| `kin_full` | Data frame from `kin2sex()` or `kin()` |
| `qx` | Probability of dying. Named list `list(f = ..., m = ...)` because `kin_full` has sex-specific kin |
| `pop` | Focal population counts (female here). A single matrix without a sex column; applies to all focal women |
| `output_year` | `NULL` computes for all years in `kin_full`; pass a vector to restrict |

**Accepted input formats for `qx` and `pop`:**

- A matrix with dimnames (ages as rownames, years as colnames)
- A long data frame with columns `year`, `age`, and the value column
- A named list `list(f = <matrix>, m = <matrix>)` for two-sex inputs

```{r run-bereavement}
result <- bereavement(
  kin_full    = kin_full,
  qx          = list(f = qx_f, m = qx_m),
  pop         = pop_mat,
  output_year = NULL          # compute for all years 1900-2018
)

glimpse(result)
```

Each row is one combination of (year, focal age, kin type). The columns are:

| Column | Description |
|--------|------------|
| `year` | Calendar year |
| `age_focal` | Age of the focal woman |
| `kin` | Relationship type code |
| `bereaved` | Expected number of bereaved women at this age = `q0 × N(age, year)` |
| `bereaved_prop` | `bereaved / N(year)` — this age's contribution to the total bereaved share |

---

## Step 6 — Inspect the probability surface

Before aggregating to counts, it is useful to look at `q0` (the per-capita probability) for a single year. This shows how bereavement risk varies across the life course for each kin type:

```{r plot-q0, fig.height=5}
result |>
  filter(year == 1980) |>
  # bereaved_prop sums to the total share; divide by pop share to get q0
  left_join(
    tibble(age_focal = 0:100, pop_share = pop_mat[, "1980"] / sum(pop_mat[, "1980"])),
    by = "age_focal"
  ) |>
  mutate(q0 = bereaved_prop / pop_share) |>
  rename_kin(sex = "f") |>
  ggplot(aes(age_focal, q0)) +
  geom_line(colour = "#e74c3c", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  facet_wrap(~ kin_label, scales = "free_y", nrow = 2) +
  labs(
    title    = "P(losing ≥1 relative this year) by focal age — Sweden 1980",
    subtitle = "Product method. Female focal, two-sex time-varying model.",
    x = "Age of focal woman", y = "Annual bereavement probability"
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.subtitle    = element_text(colour = "grey40", size = 8),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

For a 30-year-old Swedish woman in 1980, there was roughly a 1–2% chance of losing a parent in that year. Child loss is very low at that age (few children old enough to face significant mortality risk). Grandparent loss peaks in mid-life, when grandparents are typically in their 70s and 80s.

---

## Step 7 — Aggregate and summarise

`result` has one row per (year, focal age, kin). To get the **total bereaved women** per kin type per year, sum over all focal ages:

```{r aggregate}
summary_by_year <- result |>
  summarise(
    bereaved      = sum(bereaved),
    bereaved_prop = sum(bereaved_prop),   # = total bereaved / total women in year
    .by = c(year, kin)
  ) |>
  rename_kin(sex = "f")   # adds human-readable kin_label column

# Print the % of Swedish women bereaved in four selected years
summary_by_year |>
  filter(year %in% c(1950, 1970, 1990, 2010)) |>
  mutate(pct = round(bereaved_prop * 100, 2)) |>
  select(year, kin_label, pct) |>
  pivot_wider(names_from = year, values_from = pct) |>
  arrange(kin_label)
```

Note that `sum(bereaved_prop)` over focal ages equals `total_bereaved / total_population`, which is the overall per-capita bereavement rate for that kin type and year. The decline across decades for parents and grandparents reflects declining mortality in Sweden; the rise in child and grandchild losses tracks the changing population age structure.

---

## Step 8 — Plot bereaved counts by kin type

```{r barplot, fig.height=4, fig.width=12}
plot_years <- c(1950, 1970, 1990, 2010)

summary_by_year |>
  filter(year %in% plot_years) |>
  ggplot(aes(y = bereaved, x = kin_label, fill = kin_label)) +
  geom_bar(
    stat = "identity", colour = "black", linewidth = 0.25,
    show.legend = FALSE
  ) +
  scale_y_continuous(labels = scales::label_comma()) +
  coord_flip() +
  facet_wrap(~ year, nrow = 1) +
  labs(
    title    = "Swedish women who lost a relative — all-cause mortality",
    subtitle = paste0(
      "Bereaved = P(\u22651 kin loss this year) \u00d7 female population, summed over focal ages.\n",
      "Two-sex time-varying model. Female focal. Synthetic male rates. Source: DemoKin (swe_px, swe_asfr, swe_pop)."
    ),
    x = NULL,
    y = "Women who lost this relative type in the given year"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold"),
    plot.subtitle    = element_text(size = 8, colour = "grey40"),
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Parent loss accounts for the largest single category throughout the period, followed by grandparent loss. The absolute counts increase from 1950 to 2010 primarily because the total Swedish female population grew. The shift in the relative ordering of kin types over time reflects both demographic transition (falling mortality, lower fertility) and the changing age structure of the population.

---

## Input format reference

`qx` and `pop` each accept any of the following formats:

```{r input-formats, eval=FALSE}
# 1. A matrix (ages × years) — most compact for DemoKin data
qx_f_mat  # 101 × n matrix with integer rownames (ages) and colnames (years)

# 2. A long data frame (single sex)
tibble(year = 1980L, age = 0:100, qx = qx_f_mat[, "1980"])

# 3. A long data frame (two-sex, with sex column)
tibble(year = 1980L, age = 0:100, sex = "f", qx = qx_f_mat[, "1980"])

# 4. A named list of matrices (two-sex) — used in the demo above
list(f = qx_f_mat, m = qx_m_mat)
```

`qx` must be sex-specific when `kin_full` has a `sex_kin` column (two-sex mode).
`pop` does not need to be sex-specific even in two-sex mode.

---

## References

Caswell, H. (2019). The formal demography of kinship: A matrix formulation. *Demographic Research*, 41, 679–712.

Caswell, H. (2022). Formal demography of kinship IV: Two-sex models. *Demographic Research*, 47, 359–396.

Alburez-Gutierrez, D. et al. (2024). *Science Advances*, 10, eado6951. <https://doi.org/10.1126/sciadv.ado6951>
