---
title: "bereavement(): Period Bereavement Estimation with DemoKin"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  dpi = 150,
  fig.width  = 10,
  fig.height = 5
)
options(scipen = 999)
```

`bereavement()` is an R function that estimates how many people in a population lose a close relative in a given year. It is designed as a companion to the [DemoKin](https://github.com/IvanWilli/DemoKin) package, which provides the underlying kinship model.

This document walks through the full workflow from loading demographic data to producing bereavement estimates and visualisations. Along the way it explains the demographic theory behind every step, so it can serve as both a user guide and a methods reference.

---

## Background

### What is period bereavement?

A **period bereavement estimate** answers the question: *In a given calendar year, how many people lost at least one relative of type k?* This is a cross-sectional measure — a snapshot for a specific year — not a life-course cumulative probability.

Formally, for a focal individual of age *a* in year *y*, define:

```
b(a, y, k) = P(focal person loses ≥1 relative of type k during year y)
```

where *b* stands for the **bereavement probability** (not to be confused with *q_x*, the probability of dying used as input). The number of bereaved people of age *a* in year *y* is then:

```
B(a, y, k) = b(a, y, k) × N(a, y)
```

where `N(a, y)` is the number of people of age *a* alive in year *y*. Summing over all ages gives the total bereaved count for kin type *k* in year *y*.

### The kinship structure and DemoKin

To compute `q0`, we first need to know how many living relatives of each type a person of age *a* is expected to have in year *y* — and how old those relatives are. This is the **kinship structure**, a central concept in formal demography.

Caswell (2019, 2022) showed that the kinship structure can be derived from a matrix population model. Given age-specific survival probabilities and fertility rates, the model tracks expected kin counts through the recursive structure of family trees. For example: the expected number of living mothers aged *a'* for a focal person of age *a* in year *y* depends on fertility rates *a* years ago and survival rates in every year since.

The `DemoKin` package implements this model in R. Its `kin2sex()` function runs a **two-sex, time-varying** version of the model, incorporating separate male and female demographic rates that can change over time. The output is a data frame `kin_full` with one row per combination of:

| Column | Meaning |
|--------|---------|
| `year` | Calendar year |
| `age_focal` | Age of the focal person (0–100) |
| `kin` | Relationship type (see table below) |
| `age_kin` | Age of the relative (0–100) |
| `sex_kin` | Sex of the relative (`"f"` or `"m"`) |
| `living` | Expected number of living relatives with this profile |

Common kin type codes:

| Code | Relationship | Notes |
|------|-------------|-------|
| `m` | Parent (mother/father) | Expected count peaks near 1 at birth |
| `d` | Child | Grows as focal person has children |
| `gm` | Grandparent | |
| `gd` | Grandchild | |
| `s` / `os` / `ys` | Sibling (older/younger) | |
| `a` / `oa` / `ya` | Aunt or uncle | |
| `n` / `nos` / `nys` | Niece or nephew | |
| `c` / `coa` / `cya` | Cousin | |

Because `living` is derived from a deterministic matrix model, it is a **continuous-valued** expected count, not an integer. A focal person aged 30 might have an expected 0.73 grandmothers aged 68 alive — the average over the population, not a count for any specific individual.

### The product formula

Given the kinship structure, how do we compute `b(a, y, k)`?

`bereavement()` uses the **product method** (also called the "Kike method"). The only assumption it requires is that the deaths of different relatives during year *y* are **independent** given their age and sex.

Under independence, if there were exactly *K* relatives of a given age-sex profile, the probability that all of them survive year *y* would be the familiar binomial expression `(1 - q)^K`. The difficulty is that `living` — the expected kin count returned by DemoKin — is a **continuous real number**, not an integer. It is the expected value of a random variable, not an observed count.

**Why `living` is continuous and where Poisson enters.** DemoKin implements a deterministic matrix population model whose solution gives the *mean* number of relatives at each age. In the underlying stochastic process (a multitype branching process), the number of relatives of a given type converges, for large populations, to a **Poisson distribution** with mean equal to `living`. This is a consequence of the law of rare events: each relative's existence depends on a long chain of independent demographic events (surviving childbirth, surviving to reproductive age, surviving to the focal's birth year, etc.), and such chains produce Poisson-distributed counts in the limit.

**The point-mass approximation.** The product formula extends `(1 - q)^K` to continuous `living` by substituting the expected value for the random variable — treating the Poisson-distributed count as if it were fixed at its mean:

```
P(all living relatives of this age-sex profile survive) ≈ (1 - q) ^ living
                                                         = exp(living × log(1 - q))
```

This is a **point-mass (mean-field) approximation**, not an exact result. If we used the full Poisson distribution exactly, the correct expression via the probability generating function `G(z) = exp(μ(z-1))` evaluated at `z = (1-q)` would be:

```
P(all survive) [Poisson-exact] = exp(-living × q)          ← PGF of Poisson at (1-q)
P(all survive) [product method] = exp(living × log(1 - q)) ← point-mass approximation
```

These differ because `log(1-q) ≈ -q` only for small *q*. For typical demographic mortality levels — annual death probabilities in the range 0.001 to 0.05 — the two are numerically indistinguishable. For very high mortality (conflict, famine, pandemic), they diverge: the Poisson-exact formula yields a slightly higher bereavement probability than the product formula. The product method is therefore a valid and widely-used approximation for populations under normal or moderately elevated mortality, and a conservative lower bound under extreme mortality.

Taking the product over all ages *a'* and both sexes *g*:

```
P(all kin of type k survive | focal age a, year y)
  = ∏_{a', g} (1 - q(a', y, g)) ^ living(a, a', k, y, g)
```

Therefore:

```
b(a, y, k) = 1 - ∏_{a', g} (1 - q(a', y, g)) ^ living(a, a', k, y, g)
```

Ages with `living = 0` contribute `(1-q)^0 = 1` and drop out of the product automatically. This means the function only needs the full `living` surface and age-sex-specific `qx` values; it does not require any knowledge of the actual family structure of any individual.

> **Two sources of approximation error.** First, the point-mass formula replaces the Poisson kin count with its mean, which underestimates bereavement slightly when *q* is large (see above). Second, the independence assumption fails in populations where death clusters within families (conflict, epidemic, famine): deaths become positively correlated, so losses concentrate in fewer families than the formula predicts — meaning the formula both under- and over-counts depending on the quantity of interest (total bereaved vs. per-family depth of loss). For populations with such clustering, a Negative Binomial extension exists (Alburez-Gutierrez et al. 2024) but is not yet implemented here.

---

## Setup

Install `DemoKin` from GitHub if needed, then load packages and source the function:

```{r packages}
# install.packages("remotes")
# remotes::install_github("IvanWilli/DemoKin")

library(DemoKin)
library(tidyverse)

source("bereavement.R")
```

---

## Step 1 — Load demographic data

`DemoKin` ships three built-in datasets for Sweden (ages 0–100, years 1900–2018):

- `swe_px`: female annual survival probabilities `p(a, t)` — the probability of surviving from age *a* to *a+1* in year *t*
- `swe_asfr`: female age-specific fertility rates — live births per woman per year
- `swe_pop`: observed female population counts

```{r load-data}
pf <- swe_px    # female survival:  101 ages × 119 years
ff <- swe_asfr  # female fertility: 101 ages × 119 years
```

`DemoKin`'s built-in data are female only. The two-sex model requires male rates as well. Following the [DemoKin two-sex time-varying vignette](https://ivanwilli.github.io/DemoKin/articles/1_4_TwoSex_TimeVarying_Age.html), we construct synthetic male rates:

```{r male-rates}
n_ages <- nrow(pf)
n_yrs  <- ncol(pf)

# Male survival: raised to power 1.5 → uniformly lower than female (male excess mortality)
pm <- pf ^ 1.5
dimnames(pm) <- dimnames(pf)

# Male fertility: shift female ASFR 5 years to the right → later mean age at fathering
fm <- rbind(matrix(0, 5, n_yrs), ff[-((n_ages - 4):n_ages), ])
dimnames(fm) <- dimnames(ff)
```

Let's check what these rates look like for a single year:

```{r plot-rates, fig.height=4}
yr_show <- "1980"

bind_rows(
  tibble(age = 0:100, sex = "Female", component = "Survival probability",  value = pf[, yr_show]),
  tibble(age = 0:100, sex = "Male",   component = "Survival probability",  value = pm[, yr_show]),
  tibble(age = 0:100, sex = "Female", component = "Fertility rate (ASFR)", value = ff[, yr_show]),
  tibble(age = 0:100, sex = "Male",   component = "Fertility rate (ASFR)", value = fm[, yr_show])
) |>
  ggplot(aes(age, value, colour = sex)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ component, scales = "free_y") +
  scale_colour_manual(values = c(Female = "#c0392b", Male = "#2980b9")) +
  labs(
    title  = paste("Demographic rates used as model inputs — Sweden", yr_show),
    x = "Age", y = NULL, colour = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

Male survival is lower at all ages. Male fertility peaks about 5 years later than female fertility. These are the rates that drive the kinship model.

---

## Step 2 — Compute the kinship structure with `kin2sex()`

`kin2sex()` is the two-sex kinship model in `DemoKin`. It takes survival and fertility matrices for both sexes and returns, for every (focal age × year × kin type × kin age × kin sex) combination, the expected number of living and dead relatives.

Key arguments:

| Argument | Description |
|----------|------------|
| `pf`, `pm` | Female and male survival probability matrices (101 ages × *n* years) |
| `ff`, `fm` | Female and male fertility matrices (same dimensions) |
| `sex_focal` | Sex of the focal individual (`"f"` or `"m"`) |
| `time_invariant` | `FALSE` uses rates that change over time (time-varying model) |
| `birth_female` | Proportion of births that are female (typically `1/2.04 ≈ 0.49`) |
| `output_kin` | Which relationship types to compute (limits time and memory) |

The time-varying model (`time_invariant = FALSE`) is computationally intensive — it must propagate rates from every birth cohort's birth year to the present. Results are cached automatically on first run.

```{r kin2sex, eval=FALSE}
kin_out <- kin2sex(
  pf = pf, pm = pm,
  ff = ff, fm = fm,
  sex_focal      = "f",
  time_invariant = FALSE,
  birth_female   = 1 / 2.04,
  output_kin     = c("d", "gd", "gm", "m", "s")
)
kin_full <- kin_out$kin_full
```

```{r kin2sex-cache, echo=FALSE}
cache_file <- "cache/kin_full_swe_demo.rds"
if (file.exists(cache_file)) {
  kin_full <- readRDS(cache_file)
} else {
  kin_out <- kin2sex(
    pf = pf, pm = pm,
    ff = ff, fm = fm,
    sex_focal      = "f",
    time_invariant = FALSE,
    birth_female   = 1 / 2.04,
    output_kin     = c("d", "gd", "gm", "m", "s")
  )
  kin_full <- kin_out$kin_full
  dir.create("cache", showWarnings = FALSE)
  saveRDS(kin_full, cache_file)
}
```

```{r kin-dims}
nrow(kin_full)             # total rows
range(kin_full$year)       # years covered
sort(unique(kin_full$kin)) # kin types computed
```

Here is what the first few rows look like:

```{r kin-head}
kin_full |>
  filter(year == 1980, age_focal == 40, kin == "m") |>
  arrange(sex_kin, age_kin) |>
  filter(living > 0) |>
  head(10)
```

Each row describes one (age × sex) slice of a kin type for a particular focal person profile. The `living` column is the expected number of relatives with that exact age and sex still alive.

### Visualising the kinship structure

To build intuition, let's look at the expected number of living relatives by focal age for a single year:

```{r plot-kin-surface, fig.height=5}
kin_full |>
  filter(year == 1980) |>
  summarise(living = sum(living), .by = c(age_focal, kin)) |>
  rename_kin(sex = "f") |>
  ggplot(aes(age_focal, living)) +
  geom_line(colour = "#2c3e50", linewidth = 0.8) +
  facet_wrap(~ kin_label, scales = "free_y", nrow = 2) +
  labs(
    title   = "Expected number of living relatives by focal age — Sweden 1980",
    subtitle = "Two-sex time-varying model. Female focal. Living counts summed over kin age and sex.",
    x = "Age of focal woman", y = "Expected living relatives"
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.subtitle    = element_text(colour = "grey40", size = 8),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

This is the kinship structure that feeds into the bereavement calculation. For example, a 30-year-old Swedish woman in 1980 expected to have roughly one living parent, no grandchildren yet, several siblings, and a growing number of children.

---

## Step 3 — Compute the probability of dying (`qx`)

The bereavement model needs `qx`: the probability that a person aged *x* dies within the year. This is derived from the survival probabilities:

```
qx = 1 - px
```

**Terminal age problem.** `swe_px[age = 100] = 0` because everyone in the open age group (100+) is assumed to die within the year in this life table. That gives `qx = 1`, which collapses the product formula to zero for any focal person with living kin aged 100+, and triggers an error in `bereavement()`.

The fix is to approximate `qx` at age 100 using the **force of mortality** (`mx`) from age 99. The standard relationship between `px` and `mx` under a constant-hazard assumption within the age interval is `qx = 1 - exp(-mx)`, or equivalently `mx = -log(px)`. Since `px[99]` is well-defined, we compute `mx[99] = -log(px[99])` and use it as our approximation for `mx[100]`, giving `qx[100] = 1 - exp(-mx[99])`. This assumes a roughly constant force of mortality at the oldest ages, which is reasonable for period life tables. In a formal analysis, use `DemoTools::lt_single_mx()` to derive qx from mx for all ages consistently.

```{r compute-qx}
# qx = 1 - px for ages 0-99 (exact)
qx_f <- 1 - pf
qx_m <- 1 - pm

# Terminal age (row 101, age 100): px = 0 by life table convention → qx = 1.
# Approximate using mx from age 99: qx[100] = 1 - exp(-mx[99]) = 1 - px[99]
mx_f_99 <- -log(pf[100, ])   # force of mortality at age 99 (row index 100)
mx_m_99 <- -log(pm[100, ])
qx_f[101, ] <- 1 - exp(-mx_f_99)
qx_m[101, ] <- 1 - exp(-mx_m_99)

dimnames(qx_f) <- dimnames(pf)
dimnames(qx_m) <- dimnames(pm)

range(qx_f)
range(qx_m)
```

---

## Step 4 — Population data

`swe_pop` contains observed female population counts for Sweden by age and year. This is the population of focal women — we need it to convert per-capita probabilities into head counts:

```
B(a, y, k) = b(a, y, k) × N(a, y)
```

```{r pop-data}
pop_mat <- swe_pop
dimnames(pop_mat) <- list(as.character(0:100), colnames(swe_pop))

# Total Swedish women in selected years
colSums(pop_mat)[as.character(c(1950, 1970, 1990, 2010))] |>
  formatC(format = "d", big.mark = ",") |>
  print()
```

---

## Step 5 — Run `bereavement()`

`bereavement()` applies the product formula to every combination of (year, focal age, kin type) in `kin_full` and returns a tidy data frame.

**Arguments:**

| Argument | Description |
|----------|------------|
| `kin_full` | Data frame from `kin2sex()` or `kin()` |
| `qx` | Probability of dying. Named list `list(f = ..., m = ...)` in two-sex mode |
| `pop` | Population counts. Named list `list(f = ..., m = ...)` in two-sex mode; the function sums both sexes to get total population per (age, year), which is used as the focal count and denominator for `bereaved_prop` |
| `output_year` | `NULL` computes for all years in `kin_full`; pass a vector to restrict |
| `group_kin` | Named list of kin groupings, e.g. `list(nuclear = c("m", "d", "s"))`. Each group is appended to the output as an additional kin type (default `NULL`) |

**Accepted input formats for `qx` and `pop`:**

- A matrix with dimnames (ages as rownames, years as colnames)
- A long data frame with columns `year`, `age`, and the value column
- A named list `list(f = <matrix>, m = <matrix>)` for two-sex inputs

Because `kin_full` has a `sex_kin` column, both `qx` and `pop` must be named lists with `f` and `m` entries. `qx` is sex-specific because the death probability applied to each relative must match that relative's sex. `pop` is sex-specific so that the function can compute the total (female + male) population as the denominator for `bereaved_prop`.

`DemoKin` ships only female Swedish population data, so we pass `swe_pop` for both sexes here — the total population will be doubled, which is an approximation. In a real study, supply the observed male population as `m`:

```{r run-bereavement}
result <- bereavement(
  kin_full    = kin_full,
  qx          = list(f = qx_f, m = qx_m),
  pop         = list(f = pop_mat, m = pop_mat),  # female pop used for both (demo only)
  output_year = NULL
)

glimpse(result)
```

Each row is one combination of (year, focal age, kin type). The columns are:

| Column | Description |
|--------|------------|
| `year` | Calendar year |
| `age_focal` | Age of the focal woman |
| `kin` | Relationship type code |
| `bereaved` | Expected number of bereaved women at this age = `b(a,y,k) × N(a,y)` |
| `bereaved_prop` | `bereaved / N(year)` — this age's contribution to the total bereaved share |

---

## Step 6 — Inspect the bereavement probability by age

Before aggregating to counts, it is useful to look at `b(a, y, k)` — the per-capita bereavement probability — for a single year. This shows how the risk of losing each kin type varies across the life course.

`bereaved_prop` in the output is `b(a,y,k) × N(a,y) / N_total(y)` — the age-specific bereaved share. Dividing by the age's population share recovers `b(a,y,k)`:

```{r plot-p-ber, fig.height=5}
result |>
  filter(year == 1980) |>
  left_join(
    tibble(
      age_focal = 0:100,
      # total pop used inside bereavement() = pop_f + pop_m = 2 × pop_mat (demo)
      pop_share = (2 * pop_mat[, "1980"]) / sum(2 * pop_mat[, "1980"])
    ),
    by = "age_focal"
  ) |>
  mutate(p_ber = bereaved_prop / pop_share) |>   # recover b(a, y, k)
  rename_kin(sex = "f") |>
  ggplot(aes(age_focal, p_ber)) +
  geom_line(colour = "#e74c3c", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  facet_wrap(~ kin_label, scales = "free_y", nrow = 2) +
  labs(
    title    = "b(a, y, k): annual bereavement probability by focal age — Sweden 1980",
    subtitle = "Product method. Female focal, two-sex time-varying model.",
    x = "Age of focal woman", y = "b(a, y, k)"
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.subtitle    = element_text(colour = "grey40", size = 8),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  )
```

For a 30-year-old Swedish woman in 1980, there was roughly a 1–2% chance of losing a parent in that year. Child loss is very low at that age (few children old enough to face significant mortality risk). Grandparent loss peaks in mid-life, when grandparents are typically in their 70s and 80s.

---

## Step 7 — Aggregate and summarise

`result` has one row per (year, focal age, kin). To get the **total bereaved women** per kin type per year, sum over all focal ages:

```{r aggregate}
summary_by_year <- result |>
  summarise(
    bereaved      = sum(bereaved),
    bereaved_prop = sum(bereaved_prop),   # = total bereaved / total women in year
    .by = c(year, kin)
  ) |>
  rename_kin(sex = "f")   # adds human-readable kin_label column

# Print the % of Swedish women bereaved in four selected years
summary_by_year |>
  filter(year %in% c(1950, 1970, 1990, 2010)) |>
  mutate(pct = round(bereaved_prop * 100, 2)) |>
  select(year, kin_label, pct) |>
  pivot_wider(names_from = year, values_from = pct) |>
  arrange(kin_label)
```

Note that `sum(bereaved_prop)` over focal ages equals `total_bereaved / total_population`, which is the overall per-capita bereavement rate for that kin type and year. The decline across decades for parents and grandparents reflects declining mortality in Sweden; the rise in child and grandchild losses tracks the changing population age structure.

---

## Step 8 — Plot bereaved counts by kin type

```{r barplot, fig.height=4, fig.width=12}
plot_years <- c(1950, 1970, 1990, 2010)

summary_by_year |>
  filter(year %in% plot_years) |>
  ggplot(aes(y = bereaved, x = kin_label, fill = kin_label)) +
  geom_bar(
    stat = "identity", colour = "black", linewidth = 0.25,
    show.legend = FALSE
  ) +
  scale_y_continuous(labels = scales::label_comma()) +
  coord_flip() +
  facet_wrap(~ year, nrow = 1) +
  labs(
    title    = "Swedish women who lost a relative — all-cause mortality",
    subtitle = paste0(
      "Bereaved = P(\u22651 kin loss this year) \u00d7 female population, summed over focal ages.\n",
      "Two-sex time-varying model. Female focal. Synthetic male rates. Source: DemoKin (swe_px, swe_asfr, swe_pop)."
    ),
    x = NULL,
    y = "Women who lost this relative type in the given year"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold"),
    plot.subtitle    = element_text(size = 8, colour = "grey40"),
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Parent loss accounts for the largest single category throughout the period, followed by grandparent loss. The absolute counts increase from 1950 to 2010 primarily because the total Swedish female population grew. The shift in the relative ordering of kin types over time reflects both demographic transition (falling mortality, lower fertility) and the changing age structure of the population.

---

## Step 9 — Grouped kin types with `group_kin`

The `group_kin` argument lets you define composite kin categories. For each group, the function computes the probability of losing *at least one relative in the group* — still under the independence assumption — by multiplying the individual survival probabilities across all constituent kin types:

```
P(lose ≥1 kin in group | age a, year y) = 1 - ∏_{k in group} p0(a, y, k)
```

where `p0(a, y, k)` is the probability that *all* living relatives of type *k* survive the year. Group results are appended to the output alongside the individual kin types.

```{r group-kin}
result_grouped <- bereavement(
  kin_full    = kin_full,
  qx          = list(f = qx_f, m = qx_m),
  pop         = list(f = pop_mat, m = pop_mat),
  output_year = c(1950, 1970, 1990, 2010),
  group_kin   = list(
    nuclear  = c("m", "d"),      # parents + children
    ancestor = c("m", "gm")      # parents + grandparents
  )
)

result_grouped |>
  filter(kin %in% c("nuclear", "ancestor")) |>
  summarise(bereaved = sum(bereaved), bereaved_prop = sum(bereaved_prop),
            .by = c(year, kin)) |>
  mutate(pct = round(bereaved_prop * 100, 2)) |>
  select(kin, year, pct) |>
  pivot_wider(names_from = year, values_from = pct)
```

---

## Time-invariant mode

`bereavement()` also accepts `kin_full` produced by `kin(time_invariant = TRUE)` or `kin2sex(time_invariant = TRUE)`. In time-invariant mode, DemoKin sets all values in the `year` column to `NA`. The function detects this automatically and drops `year` from all join keys, so `qx` and `pop` can be passed as plain vectors or single-column matrices.

```{r time-invariant, eval=FALSE}
# Time-invariant kinship from a single year's rates
kin_ti <- kin(
  U    = swe_px[, "2010"],
  f    = swe_asfr[, "2010"],
  time_invariant = TRUE
)$kin_full

# qx and pop as plain vectors (no year dimension needed)
qx_2010  <- 1 - swe_px[, "2010"]
pop_2010 <- swe_pop[, "2010"]

# Fix terminal age
qx_2010[101] <- 1 - exp(-(-log(swe_px[100, "2010"])))

result_ti <- bereavement(
  kin_full = kin_ti,
  qx       = qx_2010,
  pop      = pop_2010
)
head(result_ti)
```

In time-invariant mode the output has `year = NA` and represents the stationary-population (stable-equivalent) bereavement probabilities implied by holding rates fixed at the supplied values.

---

## Input format reference

`qx` and `pop` each accept any of the following formats:

```{r input-formats, eval=FALSE}
# 1. A matrix (ages × years) — most compact for DemoKin data
qx_f_mat  # 101 × n matrix with integer rownames (ages) and colnames (years)

# 2. A long data frame (single sex)
tibble(year = 1980L, age = 0:100, qx = qx_f_mat[, "1980"])

# 3. A long data frame (two-sex, with sex column)
tibble(year = 1980L, age = 0:100, sex = "f", qx = qx_f_mat[, "1980"])

# 4. A named list of matrices (two-sex) — used in the demo above
list(f = qx_f_mat, m = qx_m_mat)
```

In two-sex mode (`kin_full` has a `sex_kin` column), both `qx` and `pop` must be named lists — the function enforces this. In one-sex mode, both are single matrices.

---

## Is the product method a Poisson approximation?

**Short answer:** The product method is deeply connected to Poisson reasoning, but it is
*not* identical to the standard Poisson approximation $1 - e^{-Lq}$. For low death
probabilities (baseline demography) the two are nearly indistinguishable. For large
death probabilities (conflict, pandemic, very old ages) the product formula gives
systematically *higher* bereavement — it is the more conservative estimate.
Understanding this distinction matters when comparing this function's output against the
Caswell, Verdery, and Margolis (2023) alternative, which is built on the Poisson
approximation.

---

### Background: the Poisson distribution

The **Poisson distribution** with mean $\lambda$ gives the probability of observing
exactly $k$ independent, rare events:

$$P(N = k) = \frac{e^{-\lambda} \lambda^k}{k!}, \quad k = 0, 1, 2, \ldots$$

Two properties are relevant here:

1. **Mean = variance = $\lambda$.** This is the signature of Poisson behaviour.
   Caswell (2024) shows empirically that the variance in the number of living
   *descendants* and *collateral* kin tracks the mean closely — exactly the Poisson
   fingerprint. Ancestor kin (parents, grandparents) are bounded at 2 and 4
   respectively; their variance is below their mean, making the binomial the
   appropriate distribution for them.

2. **Rare-events origin.** A Poisson arises when many independent trials each have a
   small probability of "success." Each relative's existence depends on a long chain
   of births surviving to the right ages — a process that naturally produces
   Poisson-distributed counts in large populations.

---

### Setting up the problem

Fix a focal person of age $a$ in year $y$ and a kin type $k$. For relatives of type
$k$ aged $a'$, define:

- $L = L(a, a', k, y)$: expected number of **living** kin at age $a'$, from
  `kin2sex()`. A continuous real number.
- $q = q(a', y)$: probability of dying during year $y$ at age $a'$, from the life
  table.

We want $P(\text{focal loses} \geq 1 \text{ kin of type } k \text{ in year } y)$,
which equals $1 - P(\text{all kin survive year } y)$.

There are two natural derivations depending on what we assume about the *actual*
number of kin.

---

### Case 1 — Fixed kin count: the product formula

Suppose the kin count is **fixed** at exactly $L$ (treating the mean as the actual
count). Under independence, the probability that all $L$ survive is:

$$P(\text{all survive}) = (1-q)^L = \exp(L \ln(1-q))$$

The right-hand form extends naturally to non-integer $L$, as `living` from DemoKin
is continuous. Aggregating over all kin ages $a'$ and both sexes $g$:

$$q_0(a, y, k) = 1 - \prod_{a'}\prod_{g} (1 - q(a', y, g))^{L(a, a', k, y, g)}$$

This is `bereavement()`'s formula. It is **exact** given (i) independence of deaths
and (ii) kin count equals its expected value $L$.

---

### Case 2 — Poisson-distributed kin count: the exact Poisson formula

Now suppose the actual kin count $N$ is **random**: $N \sim \text{Poisson}(L)$. Given
$N$, each individual survives independently with probability $1 - q$. The probability
that all survive, averaged over the randomness in $N$, is:

$$P(\text{all survive}) = E\!\left[(1-q)^N\right]$$

This is the **probability generating function (PGF)** of the Poisson distribution
evaluated at $z = (1-q)$.

**Deriving the PGF.** For $N \sim \text{Poisson}(\lambda)$:

$$G_N(z) = E[z^N] = \sum_{n=0}^{\infty} z^n \frac{e^{-\lambda}\lambda^n}{n!}
= e^{-\lambda} e^{z\lambda} = e^{\lambda(z-1)}$$

Setting $\lambda = L$ and $z = (1-q)$:

$$E\!\left[(1-q)^N\right] = e^{L((1-q)-1)} = e^{-Lq}$$

So under Poisson-distributed kin counts:

$$q_0 = 1 - e^{-Lq}$$

This is the **Poisson approximation** of Caswell, Verdery, and Margolis (2023, eq. 21)
and Caswell (2024, eq. 34), used for descendants and collateral kin.

---

### Comparing the two formulas

| Formula | Expression | Kin count assumption |
|---|---|---|
| **Product method** | $1 - (1-q)^L = 1 - e^{L\ln(1-q)}$ | Fixed at $L$ |
| **Poisson (Caswell et al.)** | $1 - e^{-Lq}$ | Random: $N \sim \text{Poisson}(L)$ |

They are equal only when $\ln(1-q) = -q$, which holds only at $q = 0$. For $q > 0$:

$$\ln(1-q) = -q - \frac{q^2}{2} - \frac{q^3}{3} - \cdots < -q$$

Therefore $L\ln(1-q) < -Lq$, which means $(1-q)^L < e^{-Lq}$, so:

- **Survival** probability is *lower* under the product method
- **Bereavement** probability is *higher* under the product method
- The product method is the **upper bound**

The two formulas converge as $q \to 0$ (the higher-order terms $q^2/2 + q^3/3 + \cdots$
vanish). For baseline demographic mortality ($q < 0.02$) they are numerically
indistinguishable. For conflict or epidemic mortality ($q$ can reach $0.05$–$0.30$),
the gap becomes material.

Jensen's inequality provides the underlying reason: $(1-q)^n$ is convex in $n$ for
$0 < q < 1$, so $E[(1-q)^N] \geq (1-q)^{E[N]} = (1-q)^L$. Treating $N$ as
Poisson-distributed yields a *higher* survival probability than plugging in its mean —
the product method understates survival (and overstates bereavement).

---

### Toy data: visualising the difference

The chunk below computes both formulas over the full range of $q$ for three values of
$L$ and plots them side by side. Run this after loading `tidyverse`.

```{r poisson-comparison, fig.height = 4}
expand_grid(
  q = seq(0.001, 0.99, by = 0.001),
  L = c(1, 3, 5)
) |>
  mutate(
    kike    = 1 - (1 - q)^L,
    poisson = 1 - exp(-L * q)
  ) |>
  pivot_longer(c(kike, poisson), names_to = "method", values_to = "p_bereaved") |>
  ggplot(aes(x = q, y = p_bereaved, colour = method, linetype = method)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~paste0("Expected kin  L = ", L)) +
  labs(
    x        = "Death probability  q",
    y        = "P(lose \u22651 kin)",
    title    = "Product formula vs Poisson approximation",
    subtitle = "Formulas converge as q \u2192 0; product method gives higher bereavement for large q",
    colour   = "Formula",
    linetype = "Formula"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

The table below fixes $L = 3$ (roughly the number of living siblings in mid-adulthood)
and varies $q$:

```{r poisson-table}
tibble(
  q = c(0.001, 0.010, 0.050, 0.100, 0.200, 0.300, 0.500)
) |>
  mutate(
    kike    = round(1 - (1 - q)^3, 3),
    poisson = round(1 - exp(-3 * q), 3),
    diff    = round(kike - poisson, 3)
  ) |>
  knitr::kable(
    col.names = c("q", "Product: 1-(1-q)^3", "Poisson: 1-exp(-3q)", "Difference"),
    caption   = "Bereavement probability under both formulas (L = 3 expected kin)"
  )
```

For baseline mortality (rows 1–2) the formulas agree to three decimal places. For
conflict-level mortality (rows 4–6), the product method gives 1–6 percentage points
more bereavement. At $q = 0.5$ — plausible for very old kin during intense conflict —
the gap reaches nearly 10 percentage points.

---

### Why they differ: the log approximation

The product formula uses the **exact** log: $\ln(1-q)$. The Poisson formula uses its
**first-order Taylor approximation** $\ln(1-q) \approx -q$:

$$\underbrace{1 - e^{L\ln(1-q)}}_{\text{product method}}
\approx \underbrace{1 - e^{-Lq}}_{\text{Poisson}} \quad \text{when } q \text{ is small}$$

So the product method is *not* the Poisson approximation — it is what you get by using
the exact log rather than approximating it. In low-mortality settings the approximation
is harmless. In conflict settings it is not.

---

### What about ancestor kin?

The Poisson approximation applies only to kin types that are in principle unbounded —
children, siblings, nieces, cousins. For *ancestor* kin (parents, grandparents), the
count is bounded by biology: at most 2 parents, 4 grandparents, 8 great-grandparents.
Caswell (2024) shows that the variance in ancestor counts falls well below the mean, so
the Poisson would misrepresent their distribution. The correct approximation is the
**binomial** (Caswell, Verdery, and Margolis 2023, eqs. 22–24):

$$P(\geq 1 \text{ dead parent}) = 1 - \left(1 - \frac{k_D}{2}\right)^2, \quad
P(\geq 1 \text{ dead grandparent}) = 1 - \left(1 - \frac{k_D}{4}\right)^4$$

where $k_D$ is the expected number of kin who have already died (from DemoKin's `dead`
column, differenced across years). The denominators 2 and 4 are the maximum possible
counts (Goodman, Keyfitz, and Pullum 1974).

The product method has no such complication: $(1-q)^L$ applies uniformly to all kin
types. When $L < 1$ (as for parents in old age or grandparents throughout adulthood),
the formula automatically down-weights the contribution without any distributional
assumption.

---

### Practical implications — what to watch out for

1. **Baseline demography ($q \lesssim 0.02$):** The product and Poisson formulas agree
   to within a fraction of a percentage point. Either would give essentially the same
   answer in peace-time populations.

2. **Conflict or epidemic mortality ($q \sim 0.05$–$0.30$):** The product method gives
   materially higher bereavement (5–15 percentage points for $L \approx 3$). The
   product method is more appropriate here, especially since it accepts an *externally
   supplied, cause-specific* $q_x$ directly, while the Caswell, Verdery, Margolis
   (2023) method embeds cause-specific mortality inside the kinship projection matrix
   itself.

3. **Independence is the bigger caveat.** Both formulas assume deaths of different kin
   are independent. In reality, relatives share environments — the same house, the same
   neighbourhood, the same bombardment. Positive death clustering concentrates losses
   in fewer families, so both formulas *overestimate* the share of focal persons who
   are bereaved. A Negative Binomial extension (§5.2 of `HANDOFF.txt`) can be used as
   a sensitivity check.

4. **The open question.** Caswell, Verdery, and Margolis (2023) flag that "the quality
   of these approximations is an open research question." This refers specifically to
   the Poisson/binomial approach — not to the product method, which is exact under its
   own (fixed-count, independence) assumptions.

5. **Year alignment.** The product method uses the `living` column directly for year
   $y$ and needs no offset. The Poisson method uses the *differenced* `dead` column
   (`dead` at $y+1$ minus `dead` at $y$), so it requires two consecutive years of
   kinship output. Make sure year alignment is consistent when comparing the two
   methods.

---

## References

Caswell, H. (2019). The formal demography of kinship: A matrix formulation.
*Demographic Research*, 41, 679–712.

Caswell, H. (2022). Formal demography of kinship IV: Two-sex models.
*Demographic Research*, 47, 359–396.

Caswell, H., Verdery, A. M., & Margolis, R. (2023). The formal demography of kinship
V: Kin loss, bereavement, and causes of death. *Demographic Research*, 49(41).
<https://www.demographic-research.org/articles/volume/49/41>
*[Source of the Poisson/binomial approximation formulas.]*

Caswell, H. (2024). The formal demography of kinship VI: Demographic stochasticity,
variance, and covariance in the kinship network. *bioRxiv*.
<https://doi.org/10.1101/2024.05.22.594706>
*[Variance analysis confirming Poisson for collaterals/descendants, binomial for
ancestors; eqs. 34–35 are the key formulas.]*

Goodman, L. A., Keyfitz, N., & Pullum, T. W. (1974). Family formation and the
frequency of various kinship relationships. *Theoretical Population Biology*, 5, 1–27.
*[GKP factors: maximum counts of 2 parents, 4 grandparents used in binomial formulas.]*

Song, X., Campbell, C. D., & Lee, J. Z. (2015). Ancestry matters: Patrilineage growth
and extinction. *American Sociological Review*, 80(3), 574–602.

Song, X., & Mare, R. D. (2019). Shared lifetimes, multigenerational exposure, and
educational mobility. *Demography*, 56(3), 891–916.
*[Early demographic applications of the Poisson approximation.]*

Verdery, A. M., Ryan-Claytor, C., Smith-Greenaway, E., Sarkar, N., & Livings, M.
(2024). More than 1.4 million US children have lost a family member to drug overdose.
*American Journal of Public Health*, 114(12), 1394–1397.
<https://doi.org/10.2105/AJPH.2024.307847>

Acosta, E., Alburez-Gutierrez, D., Gargiulo, M., & Torres, C. (2026). Weaponizing
kinship: A demographic analysis of bereavement in the Colombian conflict.
*Population and Development Review*.
*[Source of the product method as applied to conflict mortality.]*

Alburez-Gutierrez, D. et al. (2024). *Science Advances*, 10, eado6951.
<https://doi.org/10.1126/sciadv.ado6951>
